\begin{center}
  \textsc{Acknowledgements}
\end{center}

\noindent

I would like to express my deepest gratitude to Jan Benda, my thesis advisor,
for his support and guidance throughout the research and writing process of
this thesis. Thank you for introducing me to both programming and electric
fish, and for the opportunity to observe them in their natural habitat in
Brazil. Thank you also for entrusting me with this project and the freedom to
experiment with and venture into the field of deep learning.

I am also thankful to Nico Michiels for the many unique opportunities to learn
and grow during my studies. Thank you for the significant boost to my diving
skills, achievements that would not have been possible without my stays in
Indonesia and Corsica, which you facilitated. I am confident that these skills
will be invaluable during future field trips.

My thanks also go to Till Raab and Jan Grewe at the Neuroethology Group, whose
expertise and knowledge were crucial during my studies. Thank you, Till, for
introducing me to the analysis of electrode grid datasets and for supervising
the initial version of the chirp detector. And thank you, Jan, for always being
ready to resolve urgent programming questions, regardless of the time or day.

It was truly a pleasure working with all of you.

On a personal note, I must acknowledge the generous financial support from my
beloved family. My heartfelt gratitude goes to my mother, father, and maternal
grandmother, whose dedication and sacrifices have enabled me to pursue my
academic goals. Their belief in my potential has been a constant source of
motivation throughout this journey. Additionally, I want to thank my maternal
grandfather, who ignited my passion for nature and animals at an early age.

Among my greatest supporters has been my partner, Sofie. You brought me to
Tübingen and have been my steadfast companion throughout all of my studies.
Your unending emotional support, patience, and love have been my rock during
all academic challenges. I am deeply grateful for the countless days you spent
quizzing me with flashcards, and for your genuine interest in and thorough
understanding of my research, despite your different academic background. Our
walks in the forests around Tübingen sparked many new ideas. Thank you for
always being there, especially through the tougher times.

Lastly, I extend my gratitude to all those who were directly or indirectly
involved in the successful completion of this thesis. Your contributions have
been invaluable.
